Exploring Uncharted Territories - Predicting Activty Cliffs in Structure-Activity Landscapes

### 2012-07-24 rebuttal.tex

\DeclareGraphicsExtensions{.eps,.png} \onehalfspacing

\setlength{\parindent}{0pt} 
\setlength{\parskip}{1em} 
\usepackage{anysize}

\begin{document}

I would like to thank the reviewers for their comments on the original submissions. I have prepared
a revised manuscript that address the comments and my responses are detailed below.

@@

The text has been updated to be more specific on the development of the random forest model.

(4) The selection of training and test sets should be clearly specified in the Methods section. For
example, how were the RF models trained and evaluated for which predictions of pair-wise SALI values
are reported in Figure 2? It is hoped that the reported SALI values were not predicted for the
training data.

(5) No support is provided for the assumption that the RF models “are relatively robust even for
datasets of small size”. In fact, the author later on states several times that limited model
quality might be due to the use of small compound sets. Appropriate validation will be essential
here.

\textbf{(6) There is only limited statistical assessment/validation of the models. Statements made in the
text that R2 values are “reasonable” and the models perform “relatively well” are difficult to
reconcile. Rather, Table 2 indicates that the R2 values were rather poor in several instances.}

The text has been updated to read more quantitatively and avoid ambiguous statements

@@

Yet at the same time, the concept of an activity cliff is subjective. In many cases, it is
akin to saying \emph{``I'll know it's an activity cliff when I see it''}. For example, a pair of
molecules exhibiting a $T_c = 0.9$ and 1000x difference in acivity is obviously an activity
clif. Would a pair with $T_C = 0.85$ and 50x difference in activity not be an activity cliff? From
this point of view, I do not think it unreasonable to refer somewhat qualitatively to the
``cliffness'' of an activity cliff.

Certainly, with respect to the previous comment, a pair of molecules with $T_c = 0.29$ would likely
not be regarded as a cliff - I specifically note that such a case is not a significant cliff. To
address this aspect of SALI values, I have presented plots of dissimilarity versus activity ratio
(Figures 7 and 9) which allow one to more directly determine ``significant'' versus
``insignificant'' cliffs.

\textbf{Furthermore, there might be many instances of activity and fingerprint similarity
  relationships between pairs of compounds that might yield high pair-based SALI scores dominated by
  either the activity or similarity component, without forming a “true” activity cliff. The
  conclusion of this reviewer is that it is currently not possible to capture activity cliffs with
  any degree of certainty on the basis of the pair calculations, as reported.}

While it is true that the SALI definition itself allows for high SALI values to be generated for
pairs of molecules that would be small/insignificant cliffs (highly similar structures with small
difference in activity), the methodology does let one rank pairs of molecules and as indicated in
Figures 7 and 9, one can compare the strucural similarity and activity difference (or ratio) in 2D,
allowing one to separate ``true'' cliffs from apparent ones.

### 2012-07-25 paper.tex
some more fixes

Given the independent and dependent variables for the pairwise dataset, we can now proceed to model
development. For this study we focused on the use of random forest models\cite{Breiman:1984aa}. This
was motivated by the fact that such models tend not to overfit the training data (given a
sufficiently large number of trees) and the fact that the algorithm implicitly performs feature
selection. As a result, this allows us to forgo an explicit feature selection step and work directly
with the descriptor pool (after removal of correlated and constant descriptors). Furthermore, there
is no reason, \emph{a priori} to assume that the underlying SAR's are linear. A random forest model,
being an algorithmic approach\cite{Breiman:2001nx} (as opposed to a distributional one such as
linear regression) makes no such assumptions. We employed the implementation of random forest from
the \texttt{randomForest} package, in R 2.11.0\cite{r}. The models were built using 500 trees, and
randomly sampled the descriptor pool for a given dataset, using $\sqrt(N)$ descriptors at a time,
where $N$ is the number of descriptors in the pool. By default, the method builds individual trees
using 63\% of the dataset and tested on the remainder (the so called out-of-bag data). The final
predicted values for the entire dataset are obtained by averaging the predictions for each
observation from from all 500 trees. We also investigated the traditional approach of a
training/test set split

In this work, we employed the CDK to evaluate 109 2D and constitutional descriptors for each
individual molecule. For each dataset, we performed objective feature selection by removing
descriptors with constant or near-constant values followed by removal of descriptors that are highly
correlated with others (using an $R^2$ cutoff of 0.8). The sizes of the final descriptor pool for
each dataset are summarized in Table \ref{tab:data}. XXX

@@

\ctable[cap={}, caption={Datasets considered in this study.},
label={tab:data}]
{lp{0.4\textwidth}rrrr}
{}
{\FL
Name & Description & Endpoint & ChEMBL AID & Size & Reference
\ML
Cavalli & \raggedright{Putative hERG inhibitors (via pharmacophore modeling)} & Score &  & 30 & Ref.~\citen{Cavalli:2002aa} \NN
Costanzo & \raggedright{Thrombin inhibitors} & $\mathrm{IC_{50}}$ & 305283 & 60 &
Ref.~\citen{Costanzo:2005uq} \NN
Kalla & \raggedright{A2B adenosine receptor antagonists} & $K_i$ &
364155 & 38 & Ref.~\citen{Kalla:2006kx} \NN
Dai & \raggedright{RTK inhibitors} & $K_i$ & 429056 & 44 & Ref.~\citen{Dai:2007kl}
% Mewshaw & \raggedright{ER$\beta$ inhibitors} & $\mathrm{IC_{50}}$ &
% 306229 & 70 & \cite{Mewshaw:2005vf} \NN
\LL}

>>>>

Given the independent and dependent variables for the pairwise dataset, we can now proceed to model
development. For this study we focused on the use of random forest models\cite{Breiman:1984aa}. This
was motivated by the fact that such models tend not to overfit the training data (given a
sufficiently large number of trees) and the fact that the algorithm implicitly performs feature
selection. As a result, this allows us to forgo an explicit feature selection step and work directly
with the descriptor pool (after removal of correlated and constant descriptors). Furthermore, there
is no reason, \emph{a priori} to assume that the underlying SAR's are linear. A random forest model,
being an algorithmic approach\cite{Breiman:2001nx} (as opposed to a distributional one such as
linear regression) makes no such assumptions. We employed the implementation of random forest from
the \texttt{randomForest} package, in R 2.11.0\cite{r}. The models were built using 500 trees, and
randomly sampled the descriptor pool for a given dataset, using $\sqrt(N)$ descriptors at a time,
where $N$ is the number of descriptors in the pool. By default, the method builds individual trees
using 63\% of the dataset and tested on the remainder (the so called out-of-bag data). The final
predicted values for the entire dataset are obtained by averaging the predictions for each
observation from from all 500 trees. We also investigated the traditional approach of a
training/test set split (80\% of the dataset was assigned to the training set) and the results were
similar. As a result, the remainder of this work employed models built using the entire dataset.

In this work, we employed the CDK to evaluate 109 2D and constitutional descriptors for each
individual molecule. For each dataset, we performed objective feature selection by removing
descriptors with constant or near-constant values followed by removal of descriptors that are highly
correlated with others (using an $R^2$ cutoff of 0.8). The sizes ($N_{desc}$) of the final
descriptor pool for each dataset are summarized in Table \ref{tab:data}.

@@

\ctable[cap={}, caption={Datasets considered in this study.},
label={tab:data}]
{lp{0.3\textwidth}rrrrr}
{}
{\FL
Name & Description & Endpoint & ChEMBL AID & Size & $N_{desc}$ & Reference
\ML
Cavalli & \raggedright{Putative hERG inhibitors (via pharmacophore modeling)} & pIC$_{50}$ &  & 30 & 40 &  Ref.~\citen{Cavalli:2002aa} \NN
Costanzo & \raggedright{Thrombin inhibitors} & $\mathrm{IC_{50}}$ & 305283 & 60 & 21 & 
Ref.~\citen{Costanzo:2005uq} \NN
Kalla & \raggedright{A2B adenosine receptor antagonists} & $K_i$ &
364155 & 38 & 10 & Ref.~\citen{Kalla:2006kx} \NN
Dai & \raggedright{RTK inhibitors} & $K_i$ & 429056 & 44 & 12 & Ref.~\citen{Dai:2007kl}
% Mewshaw & \raggedright{ER$\beta$ inhibitors} & $\mathrm{IC_{50}}$ &
% 306229 & 70 & \cite{Mewshaw:2005vf} \NN
\LL}
